---
title: "Example"
author: "Gregor Steiner"
date: last-modified
editor: visual
theme: cosmo
toc: true  
number-sections: true
colorlinks: true
highlight-style: pygments
format:
  html: 
    code-fold: true
    code-tools: true
    html-math-method: katex
    self-contained: true
bibliography: references.bib
---

# Preliminaries

```{r warning=FALSE, message=FALSE}
# load packages
library(brms)
library(bayesplot)
library(ggplot2)
```


# Data

We use the well-known Lalonde dataset as in @dw_1999 to investigate the causal effect of a job training programme (**treat**) on real earnings (**re78**).


```{r}
#| tbl-cap: "Chracteristics by treatment status"

lalonde <- haven::read_dta("http://www.nber.org/~rdehejia/data/nsw_dw.dta")

within(lalonde[, -1], treat <- factor(ifelse(treat == 1, "Treated", "Control"))) |>
  gtsummary::tbl_summary(by = treat) 
```


# Causal Structure

It is reasonable to assume that the covariates $Z = (\text{age}, \text{education}, \text{black}, \text{hispanic}, \text{married}, \text{nodegree})^{\intercal}$ act as confounders, since all of them are likely to affect both the propensity to attend the job training programme as well as earnings. Let $Y$ be the real earnings and let $X$ be the treatment indicator. Then, we assume unconfoundedness, that is given covariates $Z$, the potential outcomes $\left( Y(1), Y(0) \right)$ and the treatment $X$ are independent. Under unconfoundedness, the average treatment effect (ATE) $\tau := E[Y(1) - Y(0)]$ is generally identifiable:

\begin{align*}
\tau(Z) &:= E[Y(1) - Y(0) | Z] \\ &= E[Y(1) | Z, X = 1] - E[Y(0) | Z, X = 0] \\ &= E[Y | Z, X = 1] - E[Y | Z, X = 0],
\end{align*}

which is observable and therefore
$$
\tau = E[ \tau(Z)] = \int \tau(Z) dP_Z
$$
is identified.

# Analysis

## Outcome Regression

A simple approach is regressing $Y$ on $X$ and $Z$:
$$
Y = Z \beta + \delta X + \varepsilon, \; \varepsilon \sim N(0, \Sigma)
$$
If the model is correctly specified, we get
$$
\tau(Z) = E[Y | Z, X = 1] - E[Y | Z, X = 0] = \delta,
$$

which is constant in $Z$ and therefore the ATE is $\tau = \delta$.

```{r}
fit <- lm(
  re78 ~ treat + age + education + black + hispanic + married + nodegree,
  data = lalonde
)
delta.ols <- fit$coefficients["treat"]

```


First, we estimate this using OLS and obtain a point estimate of $\widehat{\tau}_{LS} = `r round(delta.ols, 2)`$. Associated inference will be unreliable since the residuals distribution is far from a normal one, but we will ignore this issue for now and only focus on the point estimate.

Now, we estimate this in a Bayesian fashion. The result is relatively similar. The posterior distribution of $\tau$ is centered around the LS estimator $\widehat{\tau}_{LS}$.

```{r message=FALSE, warning=FALSE}
#| label: fig-delta-post
#| fig-cap: "Posterior distribution of the ATE"
fit.bayes <- brm(
  re78 ~ treat + age + education + black + hispanic + married + nodegree,
  data = lalonde,
  prior = set_prior("normal(0, 100000)", "b"),
  warmup = 1000, iter = 2000, chains = 4,
  silent = 2, refresh = 0
)


mcmc_dens(fit.bayes, pars = "b_treat") +
  labs(x = latex2exp::TeX("\\tau | (Y, X, Z)")) +
  vline_at(delta.ols, col = 2, lwd = 1) +
  theme_bw()

```









